\graphicspath{ {images/} }

\titledquestion{Position Embeddings Exploration}[6]
\label{sec:pos_enc}

Position embeddings are an important component of the Transformer architecture, allowing the model to differentiate between tokens based on their position in the sequence.
In this question, we'll explore the need for positional embeddings in Transformers and how they can be designed.

Recall that the crucial components of the Transformer architecture are the self-attention layer and the feed-forward neural network layer. 
Given an input tensor $\mathbf{X} \in \mathbb{R}^{T \times d}$, where $T$ is the sequence length and $d$ is the hidden dimension, the self-attention layer computes the following:
\begin{align*}
    \mathbf{Q} &= \mathbf{X}\mathbf{W}_Q, \quad \mathbf{K} = \mathbf{X}\mathbf{W}_K, \quad \mathbf{V} = \mathbf{X}\mathbf{W}_V \\
    \mathbf{H} &= \text{softmax}\left(\frac{\mathbf{Q}\mathbf{K}^{\top}}{\sqrt{d}}\right) \mathbf{V}
\end{align*}
where $\mathbf{W}_Q, \mathbf{W}_K, \mathbf{W}_V \in \mathbb{R}^{d \times d}$ are weight matrices, and $\mathbf{H} \in \mathbb{R}^{T \times d}$ is the output.

Next, the feed-forward layer applies the following transformation:
\begin{align*}
    \mathbf{Z} = \text{ReLU}(\mathbf{H}\mathbf{W}_1 + \mathbf{1}\cdot\mathbf{b}_1)\mathbf{W}_2 + \mathbf{1}\cdot\mathbf{b}_2
\end{align*}
where $\mathbf{W}_1, \mathbf{W}_2 \in \mathbb{R}^{d \times d}$ and $\mathbf{b}_1, \mathbf{b}_2 \in \mathbb{R}^{1\times d}$ are weights and biases; $\mathbf{1} \in \mathbb{R}^{T\times 1}$ is a vector of ones\footnote{Outer product with $\mathbf{1}$ represents broadcasting operation and makes feed forward network notations mathematically sound.}; and $\mathbf{Z} \in \mathbb{R}^{T \times d}$ is the final output.

(Note that we have omitted some details of the Transformer architecture for simplicity.)

\begin{parts}
    \part[4] \textbf{Permuting the input.}
    
    \begin{subparts}
        \subpart[3]
        Suppose we permute the input sequence $\mathbf{X}$ such that the tokens are shuffled randomly. This can be represented as multiplication by a permutation matrix $\mathbf{P} \in \mathbb{R}^{T \times T}$, i.e. $\mathbf{X}_{\text{perm}} = \mathbf{P}\mathbf{X}$. (See \href{https://en.wikipedia.org/wiki/Permutation_matrix}{Wikipedia} for a recap on permutation matrices.)

        \textbf{Show} that the output $\mathbf{Z}_{\text{perm}}$ for the permuted input $\mathbf{X}_{\text{perm}}$ will be $\mathbf{Z}_{\text{perm}} = \mathbf{P}\mathbf{Z}$.

        You are given that for any permutation matrix $\mathbf{P}$ and any matrix $\mathbf{A}$, the following hold:
        $\text{softmax}(\mathbf{P}\mathbf{A}\mathbf{P}^{\top}) = \mathbf{P}\ \text{softmax}(\mathbf{A})\ \mathbf{P}^{\top} \quad \text{and}\quad \text{ReLU}(\mathbf{P}\mathbf{A}) = \mathbf{P}\ \text{ReLU}(\mathbf{A})$.

        \ifans{
        由于输入的$\mathbf{X}$进行了一个$\mathbf{P}$变换，那么

        $$
            \mathbf{X}_{\text{new}} = \mathbf{P} \mathbf{X}
        $$

        因此可以得到

        $$
            \mathbf{Q} = \mathbf{X}_{\text{new}} \mathbf{W}_{Q} = \mathbf{P} \mathbf{X} \mathbf{W}_{Q}
        $$

        $$
            \mathbf{K} = \mathbf{X}_{\text{new}} \mathbf{W}_{K} = \mathbf{P} \mathbf{X} \mathbf{W}_{K}
        $$

        $$
            \mathbf{V} = \mathbf{X}_{\text{new}} \mathbf{W}_{V} = \mathbf{P} \mathbf{X} \mathbf{W}_{V}
        $$

        那么可以计算出$\mathbf{H}_{\text{new}}$

        \begin{align*}
            \mathbf{H}_{\text{new}} &= \text{softmax}(\frac{\mathbf{Q} \mathbf{K}^\top}{\sqrt{d}}) \mathbf{V} \\
            &= \text{softmax}(\frac{\mathbf{P} \mathbf{X} \mathbf{W}_{Q} \mathbf{W}_{k}^\top \mathbf{X}^\top \mathbf{P}^\top}{\sqrt{d}}) \mathbf{P} \mathbf{X} \mathbf{W}_{V} \\
            &= \text{softmax}(\mathbf{P} \frac{\mathbf{X} \mathbf{W}_{Q} (\mathbf{X} \mathbf{W}_{K})^\top}{\sqrt{d}} \mathbf{P}^\top) \mathbf{P} \mathbf{X} \mathbf{W}_{V} \\
            &= \mathbf{P} \text{softmax}(\frac{\mathbf{Q} \mathbf{K}^\top}{\sqrt{d}}) \mathbf{P}^\top \mathbf{P} \mathbf{X} \mathbf{W}_{V} \\
            &= \mathbf{P} \text{softmax}(\frac{\mathbf{Q} \mathbf{K}^\top}{\sqrt{d}}) \mathbf{V} \\
            &= \mathbf{P} \mathbf{H}
        \end{align*}

        得到了$\mathbf{H}_{\text{new}}$以后，即可得到$\mathbf{Z}_{\text{new}}$

        \begin{align*}
            \mathbf{Z}_{\text{new}} &= \text{ReLU}(\mathbf{H}_{\text{new}} \mathbf{W}_{1} + \mathbf{1} \cdot \mathbf{b}_{1}) \mathbf{W}_{2} + \mathbf{1} \cdot \mathbf{b}_{2} \\
            &= \text{ReLU}(\mathbf{P} \mathbf{H} \mathbf{W}_{1} + \mathbf{1} \cdot \mathbf{b}_{1}) \mathbf{W}_{2} + \mathbf{1} \cdot \mathbf{b}_{2} \\
            &= \text{ReLU}(\mathbf{P} \mathbf{H} \mathbf{W}_{1} + \mathbf{P} \cdot \mathbf{1} \cdot \mathbf{b}_{1}) \mathbf{W}_{2} + \mathbf{1} \cdot \mathbf{b}_{2} \\
            &= \text{ReLU}(\mathbf{P}(\mathbf{H} \mathbf{W}_{1} + \mathbf{1} \cdot \mathbf{b}_{1})) \mathbf{W}_{w} + \mathbf{1} \cdot \mathbf{b}_{2} \\
            &= \mathbf{P} \text{ReLU}(\mathbf{H} \mathbf{W}_{1} + \mathbf{1} \cdot \mathbf{b}_{1}) \mathbf{W}_{2} + \mathbf{P} \cdot \mathbf{1} \cdot \mathbf{b}_{2} \\
            &= \mathbf{P}(\text{ReLU}(\mathbf{H} \mathbf{W}_{1} + \mathbf{1} \cdot \mathbf{b}_{1}) \mathbf{W}_{2} + \mathbf{1} \cdot \mathbf{b}_{2}) \\
            &= \mathbf{P} \mathbf{Z}
        \end{align*}

        }

        \subpart[1] Think about the implications of the result you derived in part i. \textbf{Explain} why this property of the Transformer model could be problematic when processing text.

        \ifans{
            在输入$\mathbf{X}$的时候进行了重新排列，而在进入\text{Transformer}块前会进行位置嵌入，如果将输入的顺序打乱，会导致模型无法准确判断词之间的相对顺序，进而无法分析句子的语法结构和语义。
        }
    \end{subparts}

    \part[2] \textbf{Position embeddings} are vectors that encode the position of each token in the sequence. They are added to the input word embeddings before feeding them into the Transformer.

    One approach is to generate position embedding using a fixed function of the position and the dimension of the embedding. If the input word embeddings are $\mathbf{X} \in \mathbb{R}^{T \times d}$, the position embeddings $\Phi \in \mathbb{R}^{T \times d}$ are generated as follows:
    \begin{align*}
        \Phi_{(t, 2i)} &= \sin\left({t}/{10000^{2i/d}}\right) \\
        \Phi_{(t, 2i+1)} &= \cos\left({t}/{10000^{2i/d}}\right)
    \end{align*}
    where $t \in \{0, 1, \ldots T-1\}$ and $i \in \{0, 1, \ldots d/2-1\}$\footnote{Here $d$ is assumed even which is typically the case for most models.}.

    Specifically, the position embeddings are added to the input word embeddings:
    \begin{align*}
        \mathbf{X}_{\text{pos}} = \mathbf{X} + \Phi
    \end{align*}

    \begin{subparts}
        \subpart[1] Do you think the position embeddings will help the issue you identified in part (a)? If yes, explain how and if not, explain why not.

        \ifans{
            位置嵌入会有所帮助。位置嵌入会将每一个位置的词单独加上一个独特的位置信息，有了独特的位置信息词与词之间就可以相互判断之间的距离，从而知道句子中词的先后顺序。
        }
        \subpart[1] Can the position embeddings for two different tokens in the input sequence be the same? If yes, provide an example. If not, explain why not.

        \ifans{
            采用不同基的位置编码，当基较小的时候容易发生碰撞现象，当基较大的时候则能够为不同的位置生成更具有唯一性的编码，这是由于位置信息是正弦、余弦函数，存在周期性，当嵌入维度d较小时，也有可能发生碰撞，
            因为较高的角频率在低序维度中占据主导地位。
        }
    \end{subparts}
\end{parts}








































































































































